Time for a rethink: time sub-sampling methods in disparity-through-time analyses

### 2017-12-12 manuscript/time-slice.tex
Add appendix details

% lists - arabic in (1; (2); (3) .

@@

% NC: might move this to the NHM github account
% TG: why that? It'll be just extra hassle no?
% NC: Nah it's just a group and makes it clearer here that I'm actually doing some programming work. Takes about 30s to transfer and just means you/I need to change our remotes which again takes 30s.

% Add overview figure
%TG: This is a first draft of the figure and its caption. Tell me what you think.

  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/outlinesvg.pdf}
    \caption[Outline of the disparity-through-time pipeline]
    {1) We use ancestral character estimation to infer nodal character states; 2) we measure the pairwise Gower distance between the tip character states and nodal character states; 3) we ordinate the distance matrix using principal coordinates analysis (PCoA); 4) we time sub-sample the PCoA matrix using time bins defined by stratigraphic periods, equally sized time bins and time slices (using six methods to estimate ordination scores for branches); and finally 5) we measure disparity-through-time for each of these methods.}
    \label{overview}
  \end{figure}

\subsection{Example datasets}
\label{datasets}
To test the different time binning/slicing methods we selected four datasets: a mammal dataset from \cite{beckancient2014}, two theropod datasets from \cite{brusatte2014gradual} and \cite{bapst2016topology}, and a crinoid dataset from \cite{wright2017bayesian}.
Table \ref{table:datasets} and appendix ??? provide more details. % add appendix details
Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
These datasets are freely available with their accompanying papers (Table \ref{table:datasets}), but for reproducibility purposes we also provide the data we used on GitHub (https://github.com/nhcooper123/time-slice/data).

@@

\paragraph{Constructing distance matrices.}
To estimate the cladisto-spaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e. because some pairs of taxa share more characters in common than others, they are more likely to be similar).
For cladistic matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
However, the Gower distance cannot calculate distances when taxa have no overlapping data.
Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis} to remove pairs of taxa with no cladistic characters in common.
This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see appendix for details of which species). 
% add appendix details

@@

Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Figure in appendix, full tables in the appendix). % add details of appendix
Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
The six different time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits) show similar patterns, so we focus only of the results for one method with a punctuated model of evolution (specifically the `proximity' method), and one method with a gradual model of evolution (specifically the `gradual splits' method).
Results for all six methods can be found in the appendix. % add details of appendix

  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch.pdf}
    \caption[Relative disparity through time for four example datasets.]
    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods. 
    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).
    Results for stratigraphic ages, and for other time-slicing models, are in the appendix.}
    \label{figure:dtt1}
  \end{figure}

@@

\paragraph{Systematic differences in disparity-through-time.} 
There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods. (Table \ref{table:wilcox}, Appendix Table \ref{table:wilcox2}). 
Instead, the differences depend on the dataset and method in question.
For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different models.
Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} show different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not seen in the \cite{wright2017bayesian} dataset.
Note that for epochs, we find fewer significant differences simply because the smaller numbers of bins and slices being compared means we have low power to detect a significant difference.

@@

\paragraph{Disparity peaks}
In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time slicing rather than time binning approaches are used (Figure \ref{figure:peak1}, Appendix figure). % appendix details
This is also true for straigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, althought when using equal time bins the peaks are later than the time slicing methods, or very similar (Figure \ref{figure:peak1}, Appendix figure).
Across the three time binning methods, the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets have similar disparity peaks whichever method is used, whereas in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Figure \ref{figure:peak1}, Appendix figure).

% Plots showing the differences
  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
    \caption[Timing of peak disparity for four example datasets.]
    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method). 
    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (Table \ref{table:datasets}).
    Results for stratigraphic ages, and for other time-slicing models, are in the appendix.}
    \label{figure:peak1}
  \end{figure}

\paragraph{Effects of mass extinction events.}
Mass extinction events influence disparity in all four datasets (Figure \ref{figure:extinction1}, appendix). 
However, whether this change in disparity is significant or not depends on the dataset, and on the method used to create time sub-samples (Figure \ref{figure:extinction1}, appendix). % appendix!!!

% Big heat map table thing
  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
    \caption[Extinction four example datasets.]
    {XXX}
    \label{figure:extinction1}
  \end{figure}

@@

\section{Conclusions}

@@

\section{APPENDIX}

  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch-appendix.pdf}
    \caption[Relative disparity through time for four example datasets.]
    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, 
    % blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods. 
    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
    \label{figure:dtt2}
  \end{figure}

  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-age-appendix.pdf}
    \caption[Relative disparity through time for four example datasets.]
    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, 
    %blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods. 
    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset. 
    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
    \label{figure:dtt3}
  \end{figure}

  % Wilcoxon results table
%\begin{landscape}
  \input{tables/table_wilcox_appendix.tex} 
  \label{table:wilcox2}  
%\end{landscape}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch-appendix.pdf}
    \caption[Timing of peak disparity for four example datasets.]
    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    %Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method). 
    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
    \label{figure:peak2}
  \end{figure}

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-age-appendix.pdf}
    \caption[Timing of peak disparity for four example datasets.]
    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    %Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method). 
    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic ages. 
    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic ages in the time frame of the dataset. 
    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic ages in the time frame of the dataset. 
    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).}
    \label{figure:peak3}
  \end{figure}

>>>>

@@

% overview figure
  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/outlinesvg.pdf}
    \caption[Outline of the disparity-through-time pipeline]
    {1) We use ancestral character estimation to infer nodal character states; 2) we measure the pairwise Gower distance between the tip character states and nodal character states; 3) we ordinate the distance matrix using principal coordinates analysis (PCoA); 4) we time sub-sample the PCoA matrix using time bins defined by stratigraphic periods, equally sized time bins and time slices (using six methods to estimate ordination scores for branches); and finally 5) we measure disparity-through-time for each of these methods.}
    \label{overview}
  \end{figure}

\subsection{Example datasets}
\label{datasets}
To test the different time binning/slicing methods we selected four datasets: a mammal dataset from \cite{beckancient2014}, two theropod datasets from \cite{brusatte2014gradual} and \cite{bapst2016topology}, and a crinoid dataset from \cite{wright2017bayesian}.
Table \ref{table:datasets} and the Supporting Information Appendix S1 provide more details. 
Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
These datasets are freely available with their accompanying papers (Table \ref{table:datasets}), but for reproducibility purposes we also provide the data we used on GitHub (https://github.com/nhcooper123/time-slice/data).

@@

\paragraph{Constructing distance matrices.}
To estimate the cladisto-spaces for each of our datasets we first constructed pairwise distance matrices of length $k$, where $k$ is the total number of tips and nodes in the dataset.
We calculated the $k$$\times$$k$ distances using the Gower distance \citep{Gower71}, i.e. the Euclidean distance between two taxa divided by the number of shared characters. 
This allows us to correct for distances between two taxa that share many characters and could be closer to each other than to taxa with fewer characters in common (i.e. because some pairs of taxa share more characters in common than others, they are more likely to be similar).
For cladistic matrices, using this corrected distance is preferable to the raw Euclidean distance because of its ability to deal with discrete or/and ordinated characters as well as with missing data \citep{anderson2012using}.
However, the Gower distance cannot calculate distances when taxa have no overlapping data.
Therefore, we used the \texttt{TrimMorphDistMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis} to remove pairs of taxa with no cladistic characters in common.
This led to us removing nine taxa from the \cite{bapst2016topology} dataset, and 19 from the \cite{brusatte2014gradual} dataset, but none from the other two datasets (see Supporting Information Appendix S1 for details of which species).

@@

Disparity changes through time for each of the four datasets (Figure \ref{figure:dtt}, Supporting Information Appendix S2: Figure AX and Tables AX-X). % add details of appendix
Relative disparities tend to be lower with time binning methods, likely because these contain fewer taxa than time-slicing methods.
The six different time-slicing methods (acctran, deltran, random, proximity, equal splits and gradual splits) show similar patterns, so we focus only of the results for one method with a punctuated model of evolution (specifically the `proximity' method), and one method with a gradual model of evolution (specifically the `gradual splits' method).
Results for all six methods can be found in Supporting Information Appendix S2: Figure AX. % add details of appendix

  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-dtt-epoch.pdf}
    \caption[Relative disparity through time for four example datasets.]
    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, blue points are time slices with a punctuated model of evolution (specifically the `proximity' method), and green points are time slices with a gradual model of evolution (specifically the `gradual splits' method).
    Relative disparities (median bootstrapped disparity divided by the maximum median bootstrapped disparity for a dataset and analysis method) are presented so they can be compared across datasets/methods. 
    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    In all cases, time bin disparities are plotted at the midpoint of the bin, and error bars represent the 95\% confidence intervals around the bootstrapped median disparity.
    The four dataset names are on the first plot for each dataset (Table \ref{table:datasets}).
    Results for stratigraphic ages, and for other time-slicing models, are in the Supporting Information Appendix S2: Figure X} %appendix
    \label{figure:dtt1}
  \end{figure}

@@

\paragraph{Systematic differences in disparity-through-time.} 
There is no overall significant systematic difference among the disparities calculated using time bins and those calculated using the time-slicing methods. (Table \ref{table:wilcox}, Supporting Information Appendix S2: Table X). % appendix
Instead, the differences depend on the dataset and method in question.
For example, the \cite{brusatte2014gradual}, \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, show significant differences when using bins versus time slices defined by stratigraphy, but the \cite{beckancient2014} dataset appears robust to these different models.
Likewise, the \cite{beckancient2014}, \cite{brusatte2014gradual} and \cite{bapst2016topology} show different disparities when the number of bins or slices is the average number of stratigraphic periods, but this is not seen in the \cite{wright2017bayesian} dataset.
Note that for epochs, we find fewer significant differences simply because the smaller numbers of bins and slices being compared means we have low power to detect a significant difference.

@@

\paragraph{Disparity peaks}
In the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets, disparity peaks occur much earlier when time slicing rather than time binning approaches are used (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX). % appendix details
This is also true for straigraphic time bins in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, althought when using equal time bins the peaks are later than the time slicing methods, or very similar (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX). %appendix
Across the three time binning methods, the \cite{beckancient2014} and \cite{brusatte2014gradual} datasets have similar disparity peaks whichever method is used, whereas in the \cite{bapst2016topology} and \cite{wright2017bayesian} datasets, stratigraphic (unequal) versus equally sized time bins make a large difference to where the disparity peak occurs (Figure \ref{figure:peak1}; Supporting Information Appendix S2: Figure AX). % appendix

% Plots showing the differences
  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-peaks-epoch.pdf}
    \caption[Timing of peak disparity for four example datasets.]
    {Median bootstrapped disparities were calculated using time binning and time-slicing approaches. 
    Pink points represent time binning methods, green points are time slices with a punctuated model of evolution (specifically the `proximity' method), and blue points are time slices with a gradual model of evolution (specifically the `gradual splits' method). 
    Stratigraphy uses unequal time bins or non-equidistant slices, where the width of the bin, or the interval between slices, is equivalent to stratigraphic epochs. 
    Duration uses equal time bins or equidistant slices, where the width of the bin, or the interval between slices, is the average duration of stratigraphic epochs in the time frame of the dataset. 
    Number uses equal time bins or equidistant slices, where the number of bins, or the number of slices, is the average number of stratigraphic epochs in the time frame of the dataset. 
    For time bins there are two points indicating the maximum and minimum ages of the time bin within which peak disparities appeared.
    The four datasets are on separate rows, from top to bottom these are Beck2014, Brusatte2014, Bapst2016 and Wright2017 (Table \ref{table:datasets}).
    Results for stratigraphic ages, and for other time-slicing models, are in the Supporting Information Appendix S2: Figure AX} %appendix
    \label{figure:peak1}
  \end{figure}

\paragraph{Effects of mass extinction events.}
Mass extinction events influence disparity in all four datasets (Figure \ref{figure:extinction1}; Supporting Information Appendix S2: Figure AX). %appendix
However, whether this change in disparity is significant or not depends on the dataset, and on the method used to create time sub-samples (Figure \ref{figure:extinction1}). % appendix!!!

% Big heat map table thing
  \begin{figure}[!htbp]
    \centering
    \includegraphics[width=1\linewidth, height=1\textheight, keepaspectratio]{figures/fig-extinction-effect.pdf}
    \caption[Extinction four example datasets.]
    {XXX}
    \label{figure:extinction1}
  \end{figure}

@@

\section{Conclusions}
The way you time sub-sample your taxa has important implications for the results of disparity-through-time analyses.

\section{Data archiving statement}
Data for this study are available from \cite{dryad_n2g80,dryad_6hb7j,dryad_84t75,beckancient2014} but for ease of reproducibility can also be found on GitHub (https://github.com/nhcooper123/time-slice/data).

\section{Supporting information}
Additional Supporting Information can be found in the online version of this article:
Appendix S1: Extra details of datasets and Appendix S2: Additional tables and figures.

@@
